Geographical Python Teaching Resources: GeoPyTeR

### 2020-02-25 Geopyter.tex
Finished edits. Need to work in OER literature in 1 main paragraph and probably a few refs here and there.

@@

Although \citet{Donoho2017} traces the origins of data science back to
\citeauthor{Tukey1962}'s \textit{The future of data analysis}
(\citeyear{Tukey1962}), it is only recently that this set of previously disparate practices for working with large data sets has begun to be formalised in a way that might allow it to be taught in a university context instead of acquired on the job through `learning-by-doing'. One of the most profound practical impacts of this development has been on the ways in which developers share and execute code: although systems for combining code, commentary, and results have been around for some time (\eg R-Markdown) these were intended primarily for replicating research outputs and not for the teaching of a new technique or writing of an interactive introduction to a software library. The emergence of data science---and the overall pace of change in the tools and programming libraries that it employs---gave new impetus to the search for lightweight ways of sharing code and documentation, and an interactive browser-based technology called Jupyter was the result.

As a result, university educators in many disciplines are playing catch-up with both the methods and the tools. Geography has been acutely aware of these developments: changes in the volume and extent of spatial data available (\eg \citealp{Graham2013,gonzalez2013big,reades2016}), and in the number and range of methods employed to identify patterns in those data (\eg \citealp{Fan2016,Naik2017,Santibanez2015,Stevens2015,ArribasBel2017}), are allowing us to tackle questions thought unanswerable just a few years ago. At the same time, there is growing evidence of a `hollowing out' of once domain-specific skills \citep{Singleton2014,Singleton2016}: most of what a trained geographer used to do with specialist Geographic Information Systems (\textsc{gis}) software can now be done using free online resources (\eg Google's Fusion Tables, MapBox, Carto, ArcOnline). However, there remains unmet demand for graduates who can \emph{code} using the latest (geo)algorithms and techniques, so in this article we present a means of bringing the benefits of data science-approaches to instruction using Jupyter `notebooks' to the teaching of Python programming and geospatial analysis: Geographical Python Teaching Resources (GeoPyTer).

@@

Regardless of whether this is a revolution \citep{Wyly2014,Torrens2010} or an evolution \citep{Barnes2013,Barnes2014,ArribasBel2018}, there is no question that teaching students and researchers how to code and how to \textit{think} computationally presents new challenges \citep{Etherington2016,Muller2014,rey_09jgs} and that there is little in the way of a tradition of doing so to a high level within Geography (see surveys in \citealp{Bowlick2017,Bowlick2018}). Consequently, the majority of resources used in the classroom are being developed from scratch at each institution, often by newly-appointed lecturers and assistant professors coming from the relatively small number of elite research facilities active in this domain (see discussion in \citealp{esrc2013}). Not only has this led to the duplication of effort at multiple sites, but it has also had an enormous impact on the productivity of early career researchers, many of whom are in their first teaching post, as commonly measured by universities.

The overarching purpose of this software is therefore to address two seemingly
contradictory issues: the recognition that no two instructors teach in precisely
the same way, and the fact that few instructors have the luxury of both time and
tenure to develop compelling new course material in splendid isolation. This is 
where the Geographical Python Teaching Resource
(\textsc{g}eo\textsc{p}y\textsc{t}e\textsc{r}) comes into play: although
developed with geographers in mind, it provides a generic means by which
instructors in any discipline can selectively incorporate and re-mix programming
and conceptual content from existing Jupyter notebooks while providing their own
`gloss' for this content. Our approach therefore seeks to develop a rich community of developers and teachers who work \textit{together} on new teaching materials, who make their output entirely \emph{open} to the wider community, and whoe support instructors in quickly adapting existing materials to new delivery formats.

@@

Like many open source projects, \gp grew out of a need to scratch an itch: in our roles as teachers of (geo)computational concepts and methods we grappled on an annual basis with the demands of developing and updating instructional material with complex interdependencies. And, while the fundamental computing and analytic concepts may remain fairly stable, the field is highly-dynamic in terms of both the content (\ie what should be in the curriculum) and the code (\ie how applications should be taught). Currently, every time a widely-used library package is updated a protracted sequence of revisions ensues: some teachers, pressed for time, may resist making use of---or even alerting students to---the new features; others may leap in enthusiastically, only to discover that they hadn't fully thought through the challenges of conflicting dependencies and evolving feature sets.

We initially explored with others---faculty with whom we regularly
collaborate---the potential to `divide and conquer': we would each focus on
different topics and then combine these contributions to create a full course.
Though received with some enthusiasm, this plan did not survive contact with
reality: first, when actually faced with the uncertain promise of receiving a
full course in return for our efforts, many of us felt more secure developing
our own material; second, we had underestimated the value that we each attach to
the ability to personalise material to our own style of teaching. Since the
shared materials were being developed individually according to the individual's
own `style', any rationalisation or personalisation would have to be done by the
instructor at a later date and without support from the rest of the project.

Going back to the drawing board, we looked again at our own experiences of
successful programming and instructional efforts. In particular, we took note of
two powerful streams of thought in software and education: the importance
attached by the Free and Open Source Software (\textsc{foss}) movement to both
peer-recognition \citep{raymond_cathedral_1999} and distributed version control;
and the rise of Virtual Learning Environments (\textsc{vle}s) and of Massive
Open Online Courses (\textsc{mooc}s) with their use of rich, online learning
experiences (\eg \citealp{Trafford2011,Cabiria2012}). Reflecting on our initial
failure, we decided against trying to supply a single set of polished course
materials and instead looked for a way to provide teachers with building blocks
from which they could assemble an \textit{individualised} course suited to their
institutional and pedagogical needs.

We drew inspiration from the \textsc{foss} iPython project
\citep{perez_2007_ipython} and its transformative impact on the sharing of
integrated code, text, and multimedia resources. And we also looked to the
Python Spatial Analysis Library (\textsc{p}y\textsc{sal}) project
\cite{rey_19sea} in which contributors take ownership of application areas
through managed collaboration instead of a contributory `free-for-all'. The
subsequent emergence of the Jupyter project \citep{kluyver16} from iPython has
led to the rapid growth of a rich ecosystem of tools that support the kinds of
interaction needed to make this an effective teaching resource:
multiuser-notebooks (jupyterhub), automation of grading (nbgrader), interactive
widgets for visualisation (ipywidgets), and packages to facilitate manipulation
of the notebook itself (nbformat and notebook).

With these criteria and resources in mind, we began looking for ways to lower
the `entry costs' for faculty to both take from, and share into, the corpus of
instructional material, as well as for a way to address the
customisation/localisation challenge. \gp was our solution: a single, open hub
of teaching concepts and practical programming tasks that can be flexibly
assembled into classes, or entire courses! In the \gp approach we recognise that
teachers need to be able to develop their own classroom `story' at their own
pace (and in conformance with their institution's teaching pattern), while
supporting them with a library of up-to-date materials from which they can pick
and choose. So in much the same way that a computer application is compiled from
the contributions of many developers, \gp allows individual classes and entire
courses to be `compiled' from the contributions of many skilled teachers and
developers. As such, this positions our work within contextualised programming initiatives \citep{Guzdial2010,Lukkarinen2016} of which the CS+X approach is perhaps best-known \citep[\eg][]{Mir2017}; however, \gp represents a significant advance in this area since it not only builds in context at the level of individual modules, but at the level of a shared set of teaching resources as well!

@@

At least in our experience, the pace of technological and methodological change
is such that the gap between what instructors know and what students know is
substantial, and in many cases might even be growing. The differences within
individual student cohorts may be greater still, however, and while students
need to be able to begin their learning `journey' at some arbitrary point in the
progression from `learning to code' towards environmental, geospatial, or computational social scientist \citep{Lazer2009}. Instructors require the ability to assemble a course that meets students where they \emph{are} rather than where the instructor or institution might \emph{wish} them to be. At the same time, there is also a strong need to support more open-ended exploration during `practicals' \citep{Unwin1980}, particularly by more advanced students who may `tune out' if the means of progression is overly structured and rigid\ldots or just too slow to keep their attention.

The guiding insight behind \gp is that instruction in computation proceeds from
fundamental units of learning typically built around \textit{computing} concepts
(variables, lists/arrays, dictionaries/hashes, functions/subroutines, etc.) to
fundamental units of learning built around \textit{analytic} concepts (cluster
analysis, point patterns, spatial autocorrelation, etc.). These units must then
be assembled in a way that speaks to the student cohort and its background; we
mean this in two ways: first, that the examples used must speak to budding
geographers, political scientists, historians, etc.; and second, that it should
be possible to develop courses for different cohorts without having to start
over from scratch. In other words, how can we enable teachers to reuse many of
the building blocks employed in an advanced course for Masters students in an
introductory undergraduate course?

@@

From these constraints it was clear that our system needed to support a
`bottom-up' approach to instructional design. So although the development of a
course or class could---and probably should---start with a clear set of learning
aims and outcomes, at a certain point the instructor will be searching for a
examples with which to teach a particular concept: What \emph{is} a list or
dictionary? What is \textit{k}-means clustering? We settled on the term `Atoms'
to refer to these basic instructional units and, much like entries in the atomic
table, we felt that they could be grouped together into sets of related
concepts: the fundamentals of programming, point pattern analysis, machine
learning, etc.

\begin{wraptable}{r}{0.6\textwidth}
\centering
\caption{Overview of System Components}
\label{table:1}
\begin{tabularx}{0.6\textwidth} {sb}
 \toprule
 \textbf{Concept} & \textbf{Usage} \\
 \midrule
 Atom  & A `teachable idea' combining explanation and code for a core, typically singular, concept (\eg Variables, Lists, Projections, Robust Rescaling\ldots). \\
 Sessions  & A `teachable unit' combining \textbf{multiple atoms} for delivery in a particular learning context (\eg Beginner Lab-based Course, Intermediate Flipped Module, Online Course\ldots) \\
 Modules  & A \textbf{sequence of sessions} designed to achieve one or more pedagogical objective and/or required for credit in a particular education context. \\
\bottomrule 
\end{tabularx}
\end{wraptable}

Each atom would employ \emph{domain-specific} illustrative examples and code so as to anchor learning in problems and applications relevant to the learner. An Atom on arrays written for geographers could start by showing how these simple data structures can be used to hold data about cities (\eg name, country, population); but a subsequent unit could build on this with an an illustration of how nested data structures allow us to add city location using nested coordinate pairs (\eg latitude and longitude within an array-of-arrays). This is why managed collaboration around sets of related concepts is so important: quite aside from the fact that no \textsc{foss} project succeeds without some level of executive control, it's critical that \textit{someone} is responsible for ensuring a degree of consistency in how the units fit together within a set of closely-related topics.

The purpose of the bottom-up approach is that these units can then be flexibly assembled into Sessions: from the \textit{same ingredients} (\ie Atoms) the instructor could create two quite different modules---one for delivery in person, one for delivery online---by organising the elements in different ways. Quite simply, we don't want to have to rewrite material for each format: lecture/practical, `flipped' classroom, or online course. Furthermore, sessions designed for experts (\eg those pursuing continuing online education) might be able to `move' students through many more atoms of instruction in a single session than a similarly laid out (in terms of time) session designed for first-time programmers in an undergraduate programme. So sessions need to be able to \emph{incorporate} atoms in a way that minimises the level of effort involved in finding the `best' way to, for example, explain the concept of recursion while maximising the ability of the instructor to relate this concept to the students' practical experience (\eg by providing a `context' that is anchored in a locally relevant `story' or data).

Naturally, Sessions can then be grouped into Learning Modules (modules) that offer a coherent instructional programme over a period of weeks or months. Modules represent the highest level of abstraction in the proposed system, but they are also obviously the starting point from which instructors can organise their (remixed) atomic and sessional material into something incorporating a set of learning outcomes and a package of assessment appropriate to their students. However, our design reflects the expectation that contributions to \gp at each \emph{level} of instruction might be made by \emph{different} people: a domain expert in Spatial Bayes might be the right person to develop an Atom on the concept and its application, but not the right person to develop a module tackling advanced spatial analytic concepts where this is just one approach amongst many. Similarly, given the global diversity of delivery formats, a 10-week term in Britain enables students to cover a very different `volume' of
content from a 15-week American semester. \gp recognises and seeks to respond to
that diversity at all levels.

@@

So we are trying to design a system in which sessions and modules are
\emph{composed} out of atomic units of instruction that can be, optionally,
surrounded by the instructor's own `narrative' or supported by more locally
relevant (geo)data. We therefore want to produce a set of teaching materials
that are highly portable, easily re-used or edited, and that enable the
instructor to select only the elements from which they wish to compose their
materials. As intimated above, \gp operationalises this through the Jupyter
project and its ability to provide in-browser access to the Python
interpreter\footnote{Other languages are also possible using different kernels;
there is no reason that \gp couldn't be used to create instructional materials
in these languages as well.}, and it also takes advantage of the dominance of
the Git tool---and the GitHub service/web site---for version control as a means for tracking authorship across many versions and changes.

@@

Jupyter notebooks are written as a mix of executable code cells and
non-executable text formatted with the widely-used `markdown' syntax. Notebook
structure is provided through headers in markdown cells: a `\#' pre-pended to a
line of text is generally taken to be the title of the notebook; `\#\#' at the
start of a line provides a second level of structure (\ie level 2 headers);
`\#\#\#' indicates level 3 headers; etc. For our purposes, what's relevant is
that these headers naturally yield a semantic hierarchy that corresponds closely to the h1\ldots{}h6 model used by the HTML markup language that lies at the heart of the World Wide Web. This hierarchy allows us to `abstract out' the problem of inferring the \emph{meaning} of different sections of the notebook since the instructor does it for us through their use of headers.

@@

In order to compile Atoms into Sessions and Modules, \gp necessarily requires a compositional syntax. We've noted the conceptual mapping between markdown and \textsc{html} formatting, above but how do we select some mix of code and markdown material in one notebook to be incorporated into another? And how do we do this in a way that is both simple to express and able to resolve ambiguity? Fortunately, such a model already exists and was hinted at in Figure \ref{fig:Illustration}: Cascading Style Sheets (\textsc{css}) uses well-understood `selectors' to specify one or more elements on a web page to which a set of presentational styles should be applied.

With \textsc{css} an `h1' in a style sheet indicates that all \textsc{html} Level 1 Headers should observe the styling rules declared immediately afterwards; while `h1.important' specifies that only a Level 1 Header of the class `important' should be selected. In fact, \textsc{css} also allows for nested selectors: for some element(s) of a `larger' selection to be selected in turn. This is normally used to do things like specify mouseover behaviours for a menu: that all anchors (\ie links) that are within a division of class \texttt{menu} should act in \textit{this} way when the mouse passes over them. What is particularly elegant about this approach is that it provides a means for selecting and styling multiple pieces of content in the document in one declaration (where this is desirable) \textit{and} a means for disambiguating content with the same name but in different locations within a document's hierarchical structure (where it is not).

Conceptually, \gp reframes this in terms of the meaning imparted by the instructor to the notebook: all cells coming after a Level 1 Header are considered to be part of that element's semantic field until another Level 1 Header is encountered or the end of the document is encountered, whichever comes first. So a Level 2 Header coming `after' (or `below', if you prefer) a Level 1 Header is still part of that higher-level element's semantic field, \textit{but} we can select it uniquely within the notebook using the standard \textsc{css} form of \texttt{h1.content h2.subcontent}. And with this we have essentially repurposed \textsc{css} as a means of selecting and importing content from one notebook into another! This is illustrated in schematic form in Figure \ref{fig:structure} where the \texttt{>} imply a hierarchical relationship.

\begin{wrapfigure}{r}{0.51\textwidth}
  \centering
  \caption{Illustration of Atom Structure}
  \includegraphics[width=0.5\textwidth]{Single_Document.pdf}
  \label{fig:structure}
\end{wrapfigure}

Unfortunately, the nature of Jupyter notebooks does not allow this to happen dynamically at run-time, but it does allow something similar to happen when an instructor is `compiling' new Sessional and Module content. In short, the instructor writes whatever content they wish but, using syntax similar to the examples below, wherever they want to incorporate contributed content from \gp (or elsewhere) they have only to `include' it by specifying both a source and a selection. And this approach also works recursively: a notebook can include content from a notebook that itself includes content from another notebook\ldots

@@

Adopting the approach outlined above: we typically envision an Atom as a short notebook focussing on a core concept or method (\eg lists, object-oriented design, or spatial autocorrelation); some or all of each Atom can then be selected and imported into a Session, which is itself a notebook; and the sessions can then be selected and managed through a Module, which can be a notebook or a set of notebooks. This process is initiated by the instructor creating a blank text cell in a Jupyter notebook and writing an `include' statement. The statement should be the \textit{only} content in the cell since \gp will be replacing the cell with an unknown number of whole text and code cells from the referenced notebook.

Crucially, \texttt{include} statements can be freely intermingled with the instructor's own content, allowing the instructor to `frame' the concepts in a way that suits their teaching style but which saves them having to reinvent the wheel for each class. A Session tackling standardisation could include elements of the relevant Atom from \gp while still allowing the instructor to interject comments, observations, questions, and additional tasks to ground the learning experience in the local context (individual, institutional, etc.). To illustrate this more clearly, an Atom on Python's approach to dealing with lists could be incorporated into a longer Session as follows:

\begin{Verbatim}[fontsize=\small]
@include {
    'nb'     = 'http://geopyter.org/atoms/fundamentals/lists.ipynb',
    'select' = 'h1.Lists'
}
\end{Verbatim}

>>>>

%% For writing 'code' in LaTeX
\usepackage{textcomp}
\usepackage{listings}
\lstloadlanguages{Python,Perl,HTML}
\usepackage[dvipsnames]{xcolor}
\definecolor{LightGray}{gray}{0.95}
\lstset{% general command to set parameter(s)
	basicstyle=\small\ttfamily, % print whole listing small
	keywordstyle=\color{BrickRed}\bfseries,
	% underlined bold black keywords
	identifierstyle=\color{Fuchsia}, % nothing happens
	commentstyle=\color{Orange}\bfseries, % white comments
	stringstyle=\color{OliveGreen}, % typewriter type for strings
	showstringspaces=false} % no special string spaces
\lstset{frameround=tttt,backgroundcolor=\color{LightGray},framesep=8pt,rulecolor=\color{Gray}}

@@

Although \citet{Donoho2017} traces the origins of data science back to
\citeauthor{Tukey1962}'s \emph{The future of data analysis}
(\citeyear{Tukey1962}), it is only recently that this set of previously disparate practices for working with large data sets has begun to be formalised in a way that might allow it to be taught in a university context instead of acquired on the job through `learning-by-doing'. One of the most profound practical impacts of this development has been on the ways in which developers share and execute code: although systems for combining code, commentary, and results have been around for some time (\eg R-Markdown) these were intended primarily for replicating research outputs, with teaching and interaction as secondary considerations. The emergence of data science---and the overall pace of change in the tools and programming libraries that it employs---gave new impetus to the search for lightweight ways of sharing code and documentation, and an interactive browser-based platform called Jupyter \citep{kluyver16} was the result.

University educators in many disciplines are playing catch-up, but Geography---thanks to its long relationship with computation \citep{ArribasBel2018}---has been acutely aware of these developments: changes in the volume and extent of spatial data available (\eg \citealp{Graham2013,gonzalez2013big,reades2016}), and in the number and range of methods employed to identify patterns in those data (\eg \citealp{Fan2016,Naik2017,Santibanez2015,Stevens2015,ArribasBel2017}), are allowing us to tackle questions thought unanswerable just a few years ago. At the same time, there is evidence of a `hollowing out' of once domain-specific skills \citep{Singleton2014,Singleton2016}: most of what used to be done with specialist Geographic Information Systems (\textsc{gis}) software can now be done using free online resources (\eg Google's Fusion Tables, MapBox, Carto, ArcOnline). However, there remains unmet demand for graduates who can \emph{code} using the latest (geo)algorithms and techniques, so in this article we present one means of bringing the benefits of (spatial) data science-approaches to instruction using Jupyter `notebooks' for the teaching of the Python programming language and geospatial analysis.

@@

Regardless of whether this is a revolution \citep{Wyly2014,Torrens2010} or an evolution \citep{Barnes2013,Barnes2014}, there is no question that teaching students and researchers how to code and how to \emph{think} computationally presents new challenges \citep{Etherington2016,Muller2014,rey_09jgs} and that there is little in the way of a tradition of doing so to a high level within Geography (see surveys in \citealp{Bowlick2017,Bowlick2018}). Consequently, the majority of spatial data science and analytics resources used in the classroom are being developed from scratch at each institution, often by newly-appointed lecturers and assistant professors coming from the relatively small number of elite research facilities active in this domain (see discussion in \citealp{esrc2013}). Not only has this led to the duplication of effort at multiple sites, but it also has an enormous impact on the productivity of early career researchers, many of whom are in their first teaching post.

The overarching purpose of this software is therefore to address two seemingly contradictory issues: the recognition that no two instructors teach in precisely the same way, and the fact that few instructors have the luxury of both time and tenure to develop compelling new course material in splendid isolation. This is  where the Geographical Python Teaching Resource (\textsc{g}eo\textsc{p}y\textsc{t}e\textsc{r}) comes into play: although developed with geographers in mind, it provides a generic means by which instructors in any discipline can selectively incorporate and re-mix programming and conceptual content from existing Jupyter notebooks while providing their own `gloss' for this content. Our approach therefore seeks to develop a rich community of developers and teachers who work \emph{together} on new teaching materials, who make their output entirely \emph{open} to the wider community, and who \emph{support} instructors in quickly adapting existing materials to new delivery formats.

@@

Like many open source projects, \gp grew out of a need to scratch an itch: in our roles as teachers of (geo)computational concepts and methods we grappled on an annual basis with the demands of developing and updating instructional material with complex interdependencies. And, while the fundamental computing and analytic concepts may remain fairly stable, the field is highly-dynamic in terms of both the content (\ie what should be in the curriculum) and the code (\ie how applications should be taught). Although, in principle, there should be little interaction between curriculum and code, in practice changes to software libraries can make certain connections much harder, or much easier, to establish in the mind of the learner. For instance, until the widespread adoption of \texttt{geopandas} in Python (which itself depended on the widespread adoption of \texttt{pandas}) it was common to find that each spatial analysis library had its own data structures\footnote{A similar effect could be seen in \texttt{R} with the \texttt{tidyverse} and \texttt{sftools}.} and, consequently, students needed to be taught to work with spatial data in different ways depending on the library. Consolidation around a single approach freed up a lot of `space' in the curriculum for teaching content, not format.

In spite of this, every time a widely-used library is updated a protracted sequence of revisions ensues as teachers need to decide whether to resist making use of, or even alerting students to, the new features in order to save updating their materials, or whether to leap in enthusiastically only to struggle with conflicting dependencies and unstable feature sets. But on the basis that `many hands makes light work', we initially explored with other faculty the potential to `divide and conquer': we would each focus on different topics using the latest codebase and then combine these contributions to create a full course. Though received with some enthusiasm, this plan did not survive contact with reality: first, when actually faced with the uncertain promise of receiving a full course in return for our individual efforts, many of us felt more secure developing our own material; second, we had underestimated the value that we each attach to the ability to personalise material to our own style of teaching. Since the shared materials were being developed according to individual `style', any rationalisation or personalisation would have to be done by the instructor at a later date.

\comment{Work in OER material below}

Going back to the drawing board, we looked again at our own experiences of successful programming and instructional efforts. In particular, we took note of two powerful streams of thought in software and education: the importance attached by the Free and Open Source Software (\textsc{foss}) movement to both peer-recognition \citep{raymond_cathedral_1999} and distributed version control; and the rise of Virtual Learning Environments (\textsc{vle}s) and of Massive Open Online Courses (\textsc{mooc}s) with their use of rich, online learning experiences (\eg \citealp{Trafford2011,Cabiria2012}). Reflecting on our initial failure, we decided against trying to supply a single set of polished course materials and instead looked for a way to provide teachers with building blocks from which they could assemble an \emph{individualised} course suited to their institutional and pedagogical needs.

We drew inspiration from the original \textsc{foss} iPython project \citep{perez_2007_ipython}---expanded into the Jupyter project in 2016 \citep{kluyver16}---and its transformative impact on the sharing of integrated code, text, and multimedia resources. And we also looked to the Python Spatial Analysis Library (\textsc{p}y\textsc{sal}) project \citep{rey_19sea} in which contributors take ownership of particular application areas through managed collaboration instead of a contributory `free-for-all'. The subsequent growth of the Jupyter project has produced a rich ecosystem of tools to support the kinds of interaction needed to make this an effective teaching resource: multiuser-notebooks (\texttt{jupyterhub}), automation of grading (\texttt{nbgrader}), interactive widgets for visualisation (\texttt{ipywidgets}), and packages to facilitate manipulation of the notebook itself (\texttt{nbformat} and \texttt{notebook}).

With these criteria and resources in mind, we began looking for ways to lower the `entry costs' for faculty to both take from, and share into, a corpus of instructional material, as well as for a way to address the customisation/localisation challenge. \gp is our solution: a single, open hub of geographically-focussed teaching concepts and practical programming tasks that can be flexibly assembled into classes, or entire courses! \gp recognises that teachers need to be able to develop their own classroom `story' at their own pace (and in conformance with their institution's teaching patterns), while supporting them with a library of up-to-date materials from which they can pick and choose. So in much the same way that a computer application is compiled from the contributions of many developers, \gp allows individual classes and entire courses to be `compiled' from the contributions of many skilled teachers and developers. As such, this positions our work within contextualised programming initiatives \citep{Guzdial2010,Lukkarinen2016}, of which the CS+X approach is perhaps best-known \citep[\eg][]{Mir2017}; however, \gp represents an important advance in this area since it not only builds in context at the level of individual modules, but at the level of teaching resources as well!

@@

At least in our experience, the pace of technological and methodological change in spatial data science is such that the gap between what instructors know and what students know is substantial, and in many cases might even be growing. But the differences within individual student cohorts may be greater still, and instructors need to be able to quickly assemble and update a course that meets students where they \emph{are} rather than where the instructor or institution might \emph{wish} them to be. At the same time, there is also a strong need to support more open-ended exploration during `practicals' \citep{Unwin1980}, particularly by more advanced students who may `tune out' if progression is overly structured and rigid\ldots or just too slow to keep their attention.

The guiding insight behind \gp is that instruction in programming outside of Computer Science proceeds from fundamental units of learning typically built around \emph{computing} concepts (variables, lists/arrays, dictionaries/hashes, functions/subroutines, etc.) to fundamental units of learning built around \emph{analytic} concepts (cluster analysis, point patterns, spatial autocorrelation, etc.). These units must then be assembled in a way that speaks to the student cohort and its background; we mean this in two ways: first, that the examples used must be domain-specific in order to speak to budding geographers, political scientists, historians, etc.; and second, that it should be possible to develop courses for different types of cohorts without having to start over from scratch. In other words, how can we enable teachers to reuse many of the building blocks employed in an advanced course for Masters students in an introductory course for Undergraduates?

@@

From these constraints it was clear that our system needed to support a compositional, `bottom-up' approach to instructional design. So although the development of a course or class should obviously start out with a clear set of learning aims and outcomes, at a certain point the instructor will be searching for examples and code with which to teach a particular concept: What \emph{is} a list or dictionary? What \emph{is} \emph{k}-means clustering? We settled on the term `Atoms' to refer to these basic instructional units and, much like entries in the atomic table, we felt that they could be grouped together into sets of related concepts: the fundamentals of programming, point pattern analysis, machine learning, etc.

\begin{table}[htbp]
\centering
\caption{Overview of System Components}
\label{table:1}
\begin{tabularx}{0.6\textwidth}{sb}
 \toprule
 \textbf{Concept} & \textbf{Usage} \\
 \midrule
 Atom  & A `teachable idea' combining explanation and code for a core, typically singular, concept (\eg Variables, Lists, Projections, Robust Rescaling\ldots). \\
 Sessions  & A `teachable unit' combining \textbf{multiple atoms} for delivery in a particular learning context (\eg Beginner Lab-based Course, Intermediate Flipped Module, Online Course\ldots) \\
 Modules  & A \textbf{sequence of sessions} designed to achieve one or more pedagogical objective and/or required for credit in a particular education context. \\
\bottomrule 
\end{tabularx}
\end{table}

Each Atom would employ \emph{domain-specific} illustrative examples and code so as to anchor learning in problems and applications relevant to the learner\footnote{\gp is intended for Geography and Planning students, but could quite easily be `forked' to provide a similar set of domain-specific resources for economics, sociology, or literature students.}. An Atom could start by showing how a list can be used to hold data about cities (\eg name, country, population), and a subsequent set of `cells' (the basic `unit' of Jupyter a notebook) could build on this with an an illustration of how a list-of-lists allows us to add location as a latitude and longitude coordinate pair. We will return to some of the issues that this approach raises in \nameref{sec:engagement}, but it points to the importance of ensuring a degree of consistency in how the Atoms for a set of closely-related topics fit together.

The purpose of the bottom-up approach is that these units can then be flexibly assembled into Sessions: from the \emph{same ingredients} (\ie Atoms) the instructor could create quite different modules by organising and presenting the elements in different ways. Quite simply, we don't want to have to rewrite material for each format (\eg lecture/practical, `flipped' classroom, or distance learning), but we also need to deal with the fact that different types of scaffolding are required and that the amount of content suitable to a `class' in each of these formats might differ substantially.

Furthermore, sessions designed for experts (\eg those pursuing continuing online education) might be able to `move' students through many more Atoms of instruction in a single Session than a similarly laid out on designed for first-time programmers in an undergraduate programme. So Sessions need to be able to \emph{incorporate} Atoms in a way that minimises the level of effort involved in finding the `best' way to, for example, explain the concept of recursion while maximising the ability of the instructor to relate this concept to the students' practical experience (\eg by providing a `context' that is anchored in a locally relevant `story' or data).

Naturally, Sessions can then be grouped into Learning Modules that offer a coherent instructional programme over a period of weeks or months. Modules represent the highest level of abstraction in the proposed system, but they are also obviously the starting point from which instructors can organise their (remixed) atomic and sessional material into something incorporating a set of learning outcomes and a package of assessment appropriate to their students. However, our design reflects the expectation that contributions to \gp at each \emph{level} of instruction might be made by \emph{different} people: a domain expert in Spatial Bayes might be the right person to develop an Atom on the concept and its application, but not the right person to develop a module tackling advanced spatial analytic concepts where this is just one approach amongst many. Similarly, given the global diversity of delivery formats, a 10-week term in Britain enables students to cover a very different `volume' of content from a 15-week American semester. \gp recognises and seeks to respond to that diversity.

@@

So we are trying to design a system in which Sessions and Modules are \emph{composed} out of Atoms that can be, optionally, surrounded by the instructor's own `narrative'. We therefore want to produce a set of teaching materials that are highly portable, easily re-used or edited, and that enable the instructor to select only the elements from which they wish to compose their materials. As intimated above, \gp operationalises this through the Jupyter project and its ability to provide in-browser access to the Python interpreter\footnote{Other languages are also possible using different kernels; there is no reason that \gp couldn't be used to create instructional materials in these languages as well.}, and it also takes advantage of the dominance of the Git version control tool---and the GitHub service/web site---as a means of tracking authorship across edits.

@@

Jupyter notebooks are written as a mix of executable code cells and non-executable text formatted with the widely-used `markdown' syntax. Notebook structure is provided through headers in markdown cells: a `\#' pre-pended to a line of text is generally taken to be the title of the notebook; `\#\#' at the start of a line provides a second level of structure (\ie Level 2 headers); `\#\#\#' indicates Level 3 headers; etc. For our purposes, what's relevant is that these headers naturally yield a semantic hierarchy that corresponds closely to the h1\ldots{}h6 model used by the HTML markup language that lies at the heart of the World Wide Web. This hierarchy allows us to `abstract out' the problem of inferring the \emph{meaning} of cells in different sections of the notebook since the instructor does it for us through their use of headers.

@@

In order to assemble Atoms into Sessions and Modules, \gp necessarily requires a compositional syntax. We've noted the conceptual mapping between markdown and \textsc{html} formatting above, but how do we select some mix of code and markdown material in one notebook to be incorporated into another? And how do we do this in a way that is both simple to express and able to resolve ambiguity? Fortunately, such a model already exists and was hinted at in Figure \ref{fig:structure}: Cascading Style Sheets (\textsc{css}) uses well-understood `selectors' to specify one or more elements on a web page to which a set of presentational styles should be applied.

In \textsc{css} an `h1' in a style sheet indicates that all \textsc{html} Level 1 Headers (\eg \texttt{<h1>A Title</h1>}) should observe the styling rules declared immediately afterwards; while `h1.important' specifies that only a Level 1 Header of the class `important' should be selected (\eg \texttt{<h1 class="important">A Title</h1>}) and all other Level 1 headers ignored (\eg \texttt{<h1 class="unimportant">A Title</h1>}). In fact, \textsc{css} also allows for nested selectors in which `child' element(s) of a `parent' can be selected in turn. This is normally used to do things like specify mouseover behaviours for a menu: that all anchors (\ie links) that are within a division of class \texttt{menu} should act in \emph{this} way when the mouse passes over them (\eg \texttt{div.menu a.hover}). What is particularly elegant about \textsc{css} is that it provides a means for selecting multiple pieces of content in the document in one declaration (where this is desirable) \emph{and} a means for disambiguating content with the same name but in different locations within a document hierarchy (where it is not).

Conceptually, \gp adapts this syntax to allow us to select some or all of a Jupyter notebook using the structure imparted by the instructor: all cells coming after a Level 1 Header are considered to be part of that element's semantic field until another Level 1 Header is encountered or the end of the document is encountered, whichever comes first. And a Level 2 Header coming `after' (a `child', if you prefer) a Level 1 Header is considered part of that `parent' element's semantic field, \emph{but} we can select it uniquely within the notebook using the standard \textsc{css} form of \texttt{h1.content h2.subcontent}. This is illustrated in schematic form in Figure \ref{fig:structure}, but note that the \texttt{>} is simply make clear the hierarchical relationship. With this, we have essentially repurposed \textsc{css} as a means of selecting and importing content from one notebook into another!

\begin{figure}[htbp]
  \centering
  \caption{Illustration of a single Atom's Structure}
  \includegraphics[width=0.5\textwidth]{Single_Document.pdf}
  \label{fig:structure}
\end{figure}

Unfortunately, the nature of Jupyter notebooks does not allow this to happen dynamically at run-time, but it does allow something similar to happen when an instructor is `compiling' new Sessional and Module content. In short, the instructor writes whatever content they wish but, using syntax similar to the examples below, wherever they want to incorporate contributed content from \gp (or elsewhere) they have only to `include' it by specifying both a source and a selection. And this approach works recursively: a notebook can include content from a notebook that itself includes content from another notebook\ldots

@@

To recap, we typically envision an Atom as a short notebook focussing on a core concept or method (\eg lists, object-oriented design, or spatial autocorrelation); some or all of each Atom can then be selected and imported into a Session, which is itself a notebook; and the sessions can then be selected and managed through a Module, which can \emph{also} be a notebook or a set of notebooks. This process is initiated by the instructor creating a blank text cell in a Jupyter notebook and writing an `include' statement. The statement should be the \emph{only} content in the cell since \gp will be replacing the cell with an unknown number of whole text and code cells from the referenced notebook.

Crucially, \texttt{include} statements can be freely intermingled with the instructor's own content (as shown in Figure \ref{fig:include}, allowing the instructor to `frame' the concepts in a way that suits their teaching style but which saves them having to reinvent the wheel for each class. A Session tackling standardisation could include elements of the relevant Atom from \gp while still allowing the instructor to interject comments, observations, questions, and additional tasks to ground the learning experience in the local context (individual, institutional, etc.). To illustrate this more clearly, an Atom on Python's approach to dealing with lists could be incorporated into a longer Session as follows:

\comment{Convert into a Jupyter-like example.}

\begin{figure}
	\centering 
	\caption{Illustrative Jupyter notebook content}
	\label{fig:include}
\begin{lstlisting}[language=HTML,frame=single]
# Session 2: Lists

Module leader: Associate Prof. X
Contact information: prof.x@foo.bar.uk
\end{lstlisting}	
\begin{lstlisting}[language=Python,frame=single]
@include {
    'nb'     = 'http://geopyter.org/atoms/fundamentals/lists.ipynb',
    'select' = 'h1.Lists'
}
\end{lstlisting}

\begin{lstlisting}[language=HTML,frame=single]
## All Done?

For next week please read the following: ...
\end{lstlisting}

\end{figure}
